Safe Memory Regions for Big Data Processing

### 2016-05-27 paper/type-inference.tex
Typos fixed. Some changes from master merged.
Changes to type inference description (beginning)

\name's region type system imposes a heavy annotation burden, and the
cost of manually annotating C\# standard libraries with region types
is prohibitive. We now present our region type inference algorithm
that completely eliminates the need to write region type annotations,
except on some higher-order functions. Formally, the type inference
algorithm is an elaboration function from programs in $\absof{\FB}$
(i.e., \FB without region types, but with \C{letregion} and \C{open}
expressions, similar to the language introduced in
\S~\ref{sec:overview}) to programs in \FB.  The elaboration proceeds
in four steps. In the first step we make use of the observation that
region types are \emph{refinements} of FGJ types with region
annotations and constraints over such region annotations, and compute
polymorphic region type templates for methods and classes from their
FGJ types. The templates contain region variables ($\rho$) to denote
unknown region annotations, and predicate variables ($\varphi$) to
denote unknown constraints over such region annotations. Free region
variables are generalized in types (hence, polymorphic). Second, we
make use of the computed region type templates to elaborate
expressions by introducing region variables to denote unknown region
arguments in \C{new} expressions, method calls and function
applications. While elaborating expressions, we also build a system of
constraints that capture well-formedness requirements and subtyping
relationships between type templates that must hold (as per the static
semantics in Fig.~\ref{fb-staticsem}) for the elaboration to be valid.
Third, we lift expression elaboration and constraint generation to
methods and classes. Finally, we solve the constraints by making use
of our fixpoint constraint solving algorithm \csolvestar, which
reduces the constraint solving problem to an abduction problem in a
Herbrand constraint system, and then relies on \csolve, our abduction
solver for that domain. 
% Due to the presence of region-polymorphic
% recursion in \FB, constraints generated by the algorithm can be
% circular. More precisely, constraints generated can assume the form
% $\varphi \Leftrightarrow \phi \wedge F(\varphi)$, where $F$ is a
% non-idempotent substitution function for region variables in
% $\varphi$. In the constraint solving phase, the algorithm then relies
% on a fixpoint constraint solving algorithm called \csolvestar to solve
% the constraints and determine assignments for unknown region and
% predicate variables.

@@

The type template for a recursively defined class is computed in two
steps. First, all recursive occurances of the class among the types of
its fields are ignored and the class is templatized as if it is a
non-recursive class. Next, all the recursive occurances are
templatized with respect to the class template computed in the first
step, such that their region annotations are exactly same as the
region parameters of the class. For example, consider a generic
$\C{ListNode}\inang{\C{a}}$ class containing two fields: $\C{data}$ of
type $\C{a}$ and $\C{next}$ of type $\C{ListNode}\inang{\C{a}}$.
where $\C{a}$ is the type of the data stored in the linked list node.
To templatize the $\C{ListNode}\inang{a}$ class, we first ignore its
recursive occurance in the type of $\C{next}$ field, and templatize
the type $\C{a}$ of $\C{data}$ field as $\C{a}@\rho_0$. Based on this
type template of $\C{data}$ field, we compute the class's template as
following:
\begin{center}
$\C{ListNode}\inang{\C{a}\extends\ObjZ}
\inang{\rhoalloc_0,\rho_0\,|\,\varphi_0} \extends \ObjZ \{...\}$
\end{center}
Next, we trivially templatize the type of $\C{next}$ field in the body
of the class as $\C{ListNode}\inang{\C{a}}\inang{\rhoalloc_0,\rho_0}$.
The resulting class represents a linked list with spine in the region
$\rhoalloc_0$ and data objects in the region $\rho_0$.

% Observe that our strategy of trivially templatizing the recursive
% occurances of a class type in its definition ensures that no new
% region variables are introduced.
The templatization technique we described above for recursive class
definitions can be extended to mutually recursive definitions in a
straightforward manner, by simultaneously templatizing them. Using the
techniques outlined above, we compute region type templates for all
classes bound in the class table of the $\absof{\FB}$ program before
we proceed to elaborate expressions.

@@

A maximally weak $\phisol$ that satisfies $\isvalid{\phictxt}{\phisol
\Leftrightarrow \phicstr}$ is a constraint encoding of the smallest
graph $G$ (i.e., $\Phi(G)$) such that $G(\phictxt) \cup G$ is as
connected as $G(\phicstr)$. The problem of computing such a $G$ is
equivalent to the problem finding of finding minimum number of edges
to add to $G(\phictxt)$ such that it is as connected as $G(\phicstr)$.
Efficient algorithms to solve the later problem are known to
exist~\cite{graph}.

>>>>

\name's region type system imposes a heavy annotation burden, and the
cost of manually annotating C\# standard libraries with region types
is prohibitive. We now present our region type inference algorithm
that completely eliminates the need to write region type annotations,
except on some higher-order functions. Formally, the type inference
algorithm is an elaboration function from programs in $\absof{\FB}$
(i.e., \FB without region types, but with \C{letregion} and \C{open}
expressions, similar to the language introduced in
\S~\ref{sec:overview}) to programs in \FB.

The elaboration consists of the following steps:
\begin{enumerate}
 \item The first step consists of the following two mutually dependent steps:
   \begin{enumerate}
     \item for every class determine a set of region variable parameters ($\rho$)
       it should be parametrized over, and
     \item for every field of the class, determine its region parametrized type by
       generating as many fresh region variables as required by its class (type).
   \end{enumerate}
 \item  For every method and function type, determine its corresponding region
   type template by introducing fresh region variable parameters as required by
   the type of each parameter and return value (in addition to the allocation region
   parameter).
 \item Elaborate every \C{new} expression, method call or function application by
   introducing fresh region variable parameters (as required by the region type
   template of the called method or function).
 \item Introduce for each class and method, a predicate variable ($\varphi$) to denote
   an undetermined set of constraints over the region variable parameters of that
   class/method.
 \item We analyze each method, to generate a set of constraints that capture
   well-formedness requirements and subtyping relationships between type templates
   that must hold (as per the static semantics in Fig.~\ref{fb-staticsem}).
 \item We solve the generated set of constraints using our fixpoint constraint
   solving algorithm \csolvestar, which reduces the constraint solving problem to
   an abduction problem in a Herbrand constraint system, and then relies on \csolve,
   our abduction solver for that domain. 
 \item We finally simplify the elaborated program by coalescing region variables
   determined to be the same in the constraint solution.
\end{enumerate}

%%    This is used to replace every FGJ type in the program
%%    compute a polymorphic region type template
%%     for every class and method from their FGJ types. The templates contain
%%     region variables ($\rho$) to denote unknown region annotations, and
%%     predicate variables ($\varphi$) to denote unknown constraints over
%%     such region annotations. Free region variables are generalized in these
%%     types (which are, hence, polymorphic).
%%  
%%    Second, we make use of the computed region type templates to elaborate
%% expressions by introducing region variables to denote unknown region
%% arguments in \C{new} expressions, method calls and function
%% applications.
%% While elaborating expressions, we also build a system of
%% constraints that capture well-formedness requirements and subtyping
%% relationships between type templates that must hold (as per the static
%% semantics in Fig.~\ref{fb-staticsem}) for the elaboration to be valid.
%% Third, we lift expression elaboration and constraint generation to
%% methods and classes. Finally, we solve the constraints by making use
%% of our fixpoint constraint solving algorithm \csolvestar, which
%% reduces the constraint solving problem to an abduction problem in a
%% Herbrand constraint system, and then relies on \csolve, our abduction
%% solver for that domain.

% Due to the presence of region-polymorphic
% recursion in \FB, constraints generated by the algorithm can be
% circular. More precisely, constraints generated can assume the form
% $\varphi \Leftrightarrow \phi \wedge F(\varphi)$, where $F$ is a
% non-idempotent substitution function for region variables in
% $\varphi$. In the constraint solving phase, the algorithm then relies
% on a fixpoint constraint solving algorithm called \csolvestar to solve
% the constraints and determine assignments for unknown region and
% predicate variables.

@@

The type template for a recursively defined class is computed in two
steps. First, all recursive occurrences of the class among the types of
its fields are ignored and the class is templatized as if it is a
non-recursive class. Next, all the recursive occurrences are
templatized with respect to the class template computed in the first
step, such that their region annotations are exactly same as the
region parameters of the class. For example, consider a generic
$\C{ListNode}\inang{\C{a}}$ class containing two fields: $\C{data}$ of
type $\C{a}$ and $\C{next}$ of type $\C{ListNode}\inang{\C{a}}$.
where $\C{a}$ is the type of the data stored in the linked list node.
To templatize the $\C{ListNode}\inang{a}$ class, we first ignore its
recursive occurrence in the type of $\C{next}$ field, and templatize
the type $\C{a}$ of $\C{data}$ field as $\C{a}@\rho_0$. Based on this
type template of $\C{data}$ field, we compute the class's template as
following:
\begin{center}
$\C{ListNode}\inang{\C{a}\extends\ObjZ}
\inang{\rhoalloc_0,\rho_0\,|\,\varphi_0} \extends \ObjZ \{...\}$
\end{center}
Next, we trivially templatize the type of $\C{next}$ field in the body
of the class as $\C{ListNode}\inang{\C{a}}\inang{\rhoalloc_0,\rho_0}$.
The resulting class represents a linked list with spine in the region
$\rhoalloc_0$ and data objects in the region $\rho_0$.

% Observe that our strategy of trivially templatizing the recursive
% occurrences of a class type in its definition ensures that no new
% region variables are introduced.
The templatization technique we described above for recursive class
definitions can be extended to mutually recursive definitions in a
straightforward manner, by simultaneously templatizing them. Using the
techniques outlined above, we compute region type templates for all
classes bound in the class table of the $\absof{\FB}$ program before
we proceed to elaborate expressions.

@@

A maximally weak $\phisol$ that satisfies $\isvalid{\phictxt}{\phisol
\Leftrightarrow \phicstr}$ is a constraint encoding of the smallest
graph $G$ (i.e., $\Phi(G)$) such that $G(\phictxt) \cup G$ is as
connected as $G(\phicstr)$. The problem of computing such a $G$ is
equivalent to the problem finding of finding minimum number of edges
to add to $G(\phictxt)$ such that it is as connected as $G(\phicstr)$.
Algorithms to solve the later problem are known to
exist~\cite{siam92}.

### 2016-05-27 paper/type-system.tex
Added labels to op. semantics rules
Changes to type inference description (beginning)

Fig.~\ref{fig:fb-opsem} defines a small-step operational semantics for
\fbname via a five-place reduction relation:
\vspace*{-0.15in}
\begin{smathpar}
  \redstoo{(e,\rhomap)}{(e',\rhomap')}
\end{smathpar}
The reduction judgment should be read as following: given a set
($\rhoenv$) of regions that are currently live, and a map ($\rhomap$)
from unique identifiers of transferable regions to their current
states (Fig.~\ref{fig:region-fsm}), the expression $e$ reduces to
$e'$, while updating $\rhomap$ to $\rhomap'$. The semantics gets
``stuck'' if $e$ attempts to access an object whose allocation region
is not present in $\rhoenv$, or if $e$ tries to \C{open} a
transferable region, whose identifier is not mapped to a state by
$\rhomap$.  On the other hand, if $e$ attempts to commit an operation
on a $\RgnZ$ object that is not sanctioned by the transition
discipline in Fig.~\ref{fig:region-fsm}, then it raises an exception
value ($\invalidexn$).

@@

Furthermore, we prove the following theorem about \FB, which, in
conjunction with the type safety theorem, implies the safety of region
transfers across address spaces:
\begin{theorem}
\emph{(\textbf{Transfer Safety})}
\label{thm:fb-transfer-safety-stmt}
$\forall$ $v$, $\rhoenv$, $\rhoenv'$, $\rhomap$, $\rhomap'$, $\phicx$,
$\phicx'$, $\rgn$, $\rgn'$, such that $\rgn \in \rhoenv$, $\rgn' \in \rhoenv'$, and
$\rgn_i \notin dom(\rhomap')$, if $\hastyp{\emptyA,\rgn,
\cdot}{\C{new}\; \RgnZT{\rgn_i}}{\RgnZT{\toprgn}}$, then
$\hastyp{(\rhomap'[\rgn_i \mapsto \CLOSED],\rhoenv',\cdot,\phicx')}
{\C{new}\; \RgnZT{\rgn_i}}{\RgnZT{\toprgn}}$
\end{theorem}
The above theorem states that if a $\C{new}\; \RgnZT{\rgn_i}(v)$ value
is well-typed under one context, then it is also well-typed under
every other context, whose $\rhomap$ maps $\rgn_i$ to closed
($\CLOSED$) state.  Thus, a recipient of a transferable region only
needs to add a binding for the region to its $\rhomap$ in order to
preserve its type safety. Notably, this result could not have been
established if it were possible for a transferable region to contain
references to objects outside the region.

>>>>

Fig.~\ref{fig:fb-opsem-1} and Fig.~\ref{fig:fb-opsem-2} define a small-step
operational semantics for \fbname via a five-place reduction relation:
\vspace*{-0.15in}
\begin{smathpar}
  \redstoo{(e,\rhomap)}{(e',\rhomap')}
\end{smathpar}
The reduction judgment should be read as following: given a set
($\rhoenv$) of regions that are currently live, and a map ($\rhomap$)
from unique identifiers of transferable regions to their current
states (Fig.~\ref{fig:region-fsm}), the expression $e$ reduces to
$e'$, while updating $\rhomap$ to $\rhomap'$. The semantics gets
``stuck'' if $e$ attempts to access an object whose allocation region
is not present in $\rhoenv$, or if $e$ tries to \C{open} a
transferable region, whose identifier is not mapped to a state by
$\rhomap$.  On the other hand, if $e$ attempts to commit an operation
on a $\RgnZ$ object that is not sanctioned by the transition
discipline in Fig.~\ref{fig:region-fsm}, then it raises an exception
value ($\invalidexn$).

@@

Furthermore, we prove the following theorem about \FB, which, in
conjunction with the type safety theorem, implies the safety of region
transfers across address spaces:
\begin{theorem}
\emph{(\textbf{Transfer Safety})}
\label{thm:fb-transfer-safety-stmt}
$\forall$ $v$, $\rhoenv$, $\rhoenv'$, $\rhomap$, $\rhomap'$, $\phicx$,
$\phicx'$, $\rgn$, $\rgn'$, such that $\rgn \in \rhoenv$, $\rgn' \in \rhoenv'$, and
$\rgn_i \notin dom(\rhomap') \cup \rhoenv \cup \{\toprgn\}$, if
\[
  \hastyp{\emptyA,\rgn, \cdot}{\C{new}\; \RgnZT{\rgn_i}(v)}{\RgnZT{\toprgn}}
\]
, then
\[
  \hastyp{(\rhomap'[\rgn_i \mapsto \CLOSED],\rhoenv',\cdot,\phicx')}
{\C{new}\; \RgnZT{\rgn_i}(v)}{\RgnZT{\toprgn}}
\]
\end{theorem}
The above theorem states that if a $\C{new}\; \RgnZT{\rgn_i}(v)$ value
is well-typed under one context, then it is also well-typed under
every other context, whose $\rhomap$ maps $\rgn_i$ to closed
($\CLOSED$) state.  Thus, a recipient of a transferable region only
needs to add a binding for the region to its $\rhomap$ in order to
preserve its type safety. Notably, this result could not have been
established if it were possible for a transferable region to contain
references to objects outside the region.
